A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-02-09 paper/paper.tex
Add more stuff to the intro

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Proof by logical relation is a powerful technique that has been used
to derive metatheoretic properties of type systems, such as
consistency and parametricity. While there exists a
plethora of introductory materials about logical relation in the
context of simply typed lambda calculus and System F, a streamlined
presentation of proof by logical relation for a dependently language
is lacking in comparison. In this paper, I present a short
consistency proof for a dependently typed language that contains a
rich set of features, including an infinite and cumulative universe
hierarchy, natural numbers, and an intensional identity type. We have
fully mechanized the consistency proof using the Coq proof assistant
in under 1000 lines of code.
\end{abstract}

@@

\section{Introduction}
% Depending on its application, we care about certain metatheoretic
% properties about a type system. As a programming language, we may care
% about type soundness, which states that a well-typed never gets stuck
% during evaluation.
When a type system is used as a program logic where terms encode proofs, we
want our type system to be logically consistent, meaning
that the empty type is not inhabited.
This paper is specifically
about establishing logical consistency for a fully dependently typed
system with an infinite universe hierarchy and support for large
elimination. % The type system, presented in Section~\ref{sec:spec}, is
% most similar to Martin-Lof's predicate type theory with the minor
% difference that type conversion is based on untyped equality.

The proof of various dependently typed systems, including Martin-Lof's
type theory, has long been
established in the literature and therefore the consistency result of
this paper is not novel. Rather,
our goal is to present the proof in a form that is digestable by a
working type theorist and can be more readily mechanized in a proof
assistant. Compared to existing efforts at mechanizing logical
consistency or stronger properties such as existence of normal
form~\citep{nbeincoq},
decidable type checking~\citep{decagda}, our work is minimal since it requires very
little scaffolding and therefore results in an extremely succinct
proof of under 1000 lines of manually written Coq code for a dependent
type theory that is reasonably complete in its features.

>>>>

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Proof by logical relation is a powerful technique that has been used
to derive metatheoretic properties of type systems, such as
consistency and parametricity. While there exists a
plethora of introductory materials about logical relation in the
context of simply typed lambda calculus and System F, a streamlined
presentation of proof by logical relation for a dependently language
is lacking in comparison. In this paper, I present a short
consistency proof for a dependently typed language that contains a
rich set of features, including an infinite cumulative universe
hierarchy, booleans, and an intensional identity type. We have
fully mechanized the consistency proof using the Coq proof assistant
in under 1000 lines of code.
\end{abstract}

@@

\section{Introduction}
% Depending on its application, we care about certain metatheoretic
% properties about a type system. As a programming language, we may care
% about type soundness, which states that a well-typed never gets stuck
% during evaluation.
When a dependently type system is used as a program logic where terms
encode proofs, we want our type system to be logically consistent,
meaning that the empty type is not inhabited.

The consistency proofs of various dependently typed systems, including Martin-Lof's
type theory and the Calculus of Constructions, have long been
available in the literature. In particular, recent works such as \citet{nbeincoq},
\citet{decagda}, and \citet{martin-lof-a-la-coq} mechanize stronger
results related to the decidability of type conversion or type
checking of dependently typed systems, from which consistency can be
derived as a corollary. The underlying technique of the forementioned
works is proof by logical relation,
which involves interpreting types as reducibility predicates,
representing sets of terms satisfying certain properties with respect
to the reduction relation. While the proof technique and the
consistency result for dependent types are both well-established,
there is a severe lack of rigorous and accessible material that shows
how proof by logical relation can be applied to dependently typed
systems.
Introductory materials about logical relations such as
\citet{skorstengaard2019introduction}, \citet{harper2016practical}
start from logical relations for simply typed languages, and
eventually extend the technique to build a relational model for System
F in order to derive
parametricity. \citet{harperkripke} gives a gentle introduction on how
logical relations for closed terms can be extended to talk about open
terms so we can derive properties such as normalization for well-typed
open terms. In short, the introductory texts about logical relations
cover systems and properties with varying degrees of complexity, from
simply typed to polymorphicly typed, logical predicate to logical
equivalence, closed terms to open terms. The glaring gap here is the
lack of fully dependently typed systems where computations may appear
at the type-level. In fact, it is far from obvious why proof by
logical relation is even applicable to dependent types, since the type
may very well-be a computation that is yet to be evaluated.
While it is assuring that proof by logical relations for dependent
types is available in mechanized forms, % the key to address the
% complexities of dependent types is obscured in
\citet{nbeincoq,decagda,martin-lof-a-la-coq} all involve relational,
Kripke-style models that obscure the underlying technique of applying
proof by logical relation to dependent types. In fact, the
complexivity is self-evident from the size of their code base. For
example, \citet{nbeincoq} involves 22 thousand lines of Coq code.

The goal of this paper is to give a tutorial on proof by logical relation for
dependent types in a simple and digestable format. % The language specification will
% be given in Section~\ref{sec:spec}.
The system we present is small but not minimal. We want to cover
features that would require more interesting treatments under
dependent types without introducing complexities that would have
existed in its simply typed variant. We choose boolean
types over natural numbers as our base type for simplicity, but include an
intensional identity type in our type system as the interpretation
of identity types requires extra side conditions that are not present
in other types. Unlike \citet{nbeincoq,decagda,martin-lof-a-la-coq}
but similar to \citet{anand2014towards}, we include an infinite
hierarchy of universes since we want to not only have at least one
predicative universe level to support type-level computations, but
also avoid the unnecessary code duplication pointed out by
\citet{nbeincoq}.
We choose consistency as the property of interest since it is the one
of the easiest metatheoretic property that requires proof by logical
relation.

Our consistency proof is short and fully mechanized. The proof scripts
involve less than 1000 lines of Coq code. Furthermore, the mechanized
proof closely corresponds to the informal

% A naive attempt at proving consistency through induction over the
% typing derivation would fail since the inductive hypothesis is not
% strong enough to derive the consistency result. Instead, one typically
% relies on the technique referred to as proof by logical relation to
% interpret types as reducibility predicates to strengthen the inductive
% hypothesis.

This paper is specifically
about establishing logical consistency for a fully dependently typed
system with an infinite universe hierarchy and support for large
elimination . % The type system, presented in Section~\ref{sec:spec}, is
% most similar to Martin-Lof's predicate type theory with the minor
% difference that type conversion is based on untyped equality.

Rather,
our goal is to present the proof in a form that is digestable by a
working type theorist and can be more readily mechanized in a proof
assistant. Compared to existing efforts at mechanizing logical
consistency or stronger properties such as existence of normal
form~\citep{nbeincoq},
decidable type checking~\citep{decagda}, our work is minimal since it requires very
little scaffolding and therefore results in an extremely succinct
proof of under 1000 lines of manually written Coq code for a dependent
type theory that is reasonably complete in its features.
